One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-05 paper/appendix.tex
appendix and minor updates

\section{Modification to vertical structure by self-gravity}\label{vertsg_mod}
We describe a simple procedure to set up the vertical structure of a locally
isothermal, self-gravitating disc. We imagine setting up a non-self-gravitating disc,
then slowly switch on the vertical force due to self-gravity. We expect the midplane
density to increase at the expense of gas density higher in the atmosphere. It is assumed
that the temperature profile remains unchanged.

Vertical hydrostatic equilibrium between
gas pressure, stellar gravity and self-gravity reads
\begin{align}
c_\mathrm{iso}^2(R)\frac{\p\ln{\rho}}{\p z} = -\frac{\p\Phi_*}{\p z} - \frac{\p\Phi}{\p z}. 
\end{align}
Assuming a smooth radial density profile, we use the plane-parallel atmosphere
approximation for the disc potential, i.e.
\begin{align}
\frac{\p^2\Phi}{\p z^2} = 4\pi G \rho. 
\end{align}
Next, we write the density field as
\begin{align}
\rho(R,z) = \rho_N(R, z) \times \beta(z; R)
\end{align}
where $\rho_N$ is the density field corresponding to the non-self-gravitating disc:
\begin{align}
&c_\mathrm{iso}^2(R)\frac{\p\ln{\rho_N}}{\p z} = -\frac{\p\Phi_*}{\p z},\\
&\rho_N = \frac{\Sigma(R)}{\sqrt{2\pi}H(R)}\exp{(-z^2/2H^2)} \notag\\
&\phantom{\rho_N}\equiv \rho_{N0}(R)\exp{(-z^2/2H^2)},
\end{align}
where $\rho_{N0}=\rho_N(R,z=0)$ is the midplane density. The explicit
expression for $\rho_N$ above assumes a thin disc. The function
$\beta$ describes the modification to the local density in order to be
consistent with self-gravity. By construction, its governing equation
is 
\begin{align}
c_\mathrm{iso}^2(R)\frac{\p^2\ln{\beta}}{\p z^2} = -4\pi G\rho_N\beta. 
\end{align}
Let 
\begin{align}
& \chi  = \ln\beta - z^2/2H^2 ,\\
& \xi= \left(\frac{4\pi G \rho_{N0} }{c_\mathrm{iso}^2}\right)^{1/2} z,
\end{align}
then the governing equation can be written in dimensionless form
\begin{align}
&\frac{\p^2 \chi}{\p\xi^2} = -K -\exp{\chi}\label{vertsg_eqn},\\
& K \equiv \frac{c_\mathrm{iso}^2}{4\pi G \rho_{N0} H^2}. \notag
\end{align}
Note that $K(R)$ is proportional to the local Keplerian Toomre
parameter. Eq. \ref{vertsg_eqn} can be further reduced to a first
order differential equation, but this is unnecessary because we
pursue a numerical solution at the end. Appropriate 
boundary conditions are
\begin{align}
&\chi(z=0)  =  \ln\beta_0, \\
&\left.\frac{\p \chi}{\p \xi} \right|_{z=0} = 0.
\end{align}
$\beta_0(R)$ is the midplane density enhancement. To determine its value, we impose
the surface density before and after modification by self-gravity to
remain the same. Then we require 
\begin{align}\label{beta0_eqn}
&F(\beta_0) \equiv \sqrt{\frac{2K}{\pi}}\int_0^{ n /\sqrt{K} } 
\exp{\chi(\xi;\beta_0)}d\xi - \erf{\left(\frac{n}{\sqrt{2}}\right)} \notag\\
&\phantom{F(\beta_0)} = 0,  
\end{align}
where $n$ is the number of scale-heights of the non-self-gravitating
disc we originally considered. At a given cylindrical radius $R$, we
solve Eq. \ref{beta0_eqn} using Newton-Raphson iteration. Each
iteration involves integrating the governing ODE for $\chi$
(Eq. \ref{vertsg_eqn}). At the end of the iteration, we have $\beta(z;
R)$ and the midplane enhancement $\beta_0(R)$.

We comment that the procedure outlined above can be extended to
polytropic discs. In this case, there is an additional unknown --- 
the new disc thickness after adjustment by self-gravity and an
additional constraint --- the density should vanish at the new disc
surface.

>>>>

\section{The background torque density in a three-dimensional disc with
  a fixed temperature profile}\label{tbg_deriv}
We give a brief derivation of the angular momentum exchange between
linear perturbations and the background disc. We consider a
three-dimensional disc in which the equilibrium pressure and density
are related by 
\begin{align}\label{iso_cond}
  p = c_s^2(R,z)f(\rho),
\end{align} 
where $f(\rho)$ is an arbitrary function of $\rho$ with dimensions of
mass per unit volume, and $c_s$ is a prescribed function of
position with dimensions of velocity squared. The equilibrium disc
satisfies 
\begin{align}
  R\Omega^2(R,z) &= \frac{1}{\rho}\frac{\p p}{d R} +
  \frac{\p\Phi_\mathrm{tot}}{\p R}, \\
  0 &= \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_\mathrm{tot}}{\p
    z}. 
\end{align}
Note that, in general, the equilibrium rotation $\Omega$ depends on
$R$ and $z$.

We begin with the linearised equation of motion in terms of the
Lagragian displacement $\bm{\xi}$ as given by \cite{lin93b} but with an
additional potential perturbation, 
\begin{align}\label{lagragian_pert}
  &\frac{D^2\bm{\xi}}{Dt^2} +
  2\Omega\hat{\bm{z}}\times\frac{D\bm{\xi}}{Dt}  \notag \\ &= -
  \frac{\nabla \delta p }{\rho} + \frac{\delta\rho}{\rho^2}\nabla\rho  
  -\nabla\delta\Phi_d - R
  \hat{\bm{R}}\left(\bm{\xi}\cdot\nabla\Omega^2\right) \notag \\
  & = -\nabla\left(\frac{\delta p}{\rho} + \delta \Phi_d\right) -
  \frac{\delta p}{\rho}\frac{\nabla\rho}{\rho} +
  \frac{\delta\rho}{\rho}\frac{\nabla p}{\rho} - - R
  \hat{\bm{R}}\left(\bm{\xi}\cdot\nabla\Omega^2\right),
\end{align}
where $D/Dt \equiv \p_t + \ii m \Omega$ for perturbations with
azimuthal dependence in the form $\exp\left(\ii m \phi\right)$, and  
the $\delta$ quantities denote Eulerian perturbations.

As explained in \cite{lin11b}, a conservation law for the angular
momentum of the perturbation may be obtained by taking the dot product
between Eq. \ref{lagragian_pert} and $(-m/2)\rho\bm{\xi}^*$, then
taking the imaginary part afterwards. The left hand side bcomes the
angluar momentum density. The first term on the right hand side (RHS)
becomes 
\begin{align}\label{angflux1}
  &-\frac{m}{2}\imag\left[-\rho\bm{\xi}^*\cdot\nabla\left(\frac{\delta p}{\rho} + \delta
    \Phi_d\right)\right] \notag\\ 
&= \frac{m}{2}\imag\left\{\nabla\cdot\left[\rho\bm{\xi}^*\left(\frac{\delta p}{\rho} + \delta
    \Phi_d \right) + \frac{1}{4\pi
    G}\delta\Phi_d\nabla\delta\Phi_d^*\right]\right\} \notag\\
&+ \frac{m}{2}\imag\left(\delta\rho^*\frac{\delta p}{\rho}\right),
\end{align}
where $\delta\rho = - \nabla\cdot\left(\rho\bm{\xi}\right)$ and
$\nabla^2\delta\Phi_d = 4\pi G \delta \rho$ have been used. The first
term on the RHS of Eq. \ref{angflux1} is (minus) the 
angular momentum flux. The second term on RHS of Eq. \ref{angflux1}, 
together with the remaining terms on the RHS of 
Eq. \ref{lagragian_pert} constitutes the background torque. That is, 
\begin{align}\label{tbg_3d}
  T_\mathrm{BG} = \frac{m}{2}\imag\left[
    \frac{\delta p}{\rho} \Delta\rho^* -
%\left(\delta\rho^* + \bm{\xi}^*\cdot\nabla\rho\right) - 
    \frac{\delta\rho}{\rho}\bm{\xi}^*\cdot\nabla p  
    + \rho \xi_R^*\xi_z\frac{\p\left(R\Omega^2\right)}{\p z}\right],
\end{align}
where $\Delta\rho = \delta\rho + \bm{\xi}\cdot\nabla\rho$ is the
Lagragian density perturbation.

So far we have not invoked an energy equation. For 
adiabatic perturbations $T_\mathrm{BG}$ is zero
\citep{lin93b}. However, if we impose the equilibrium relation
Eq. \ref{iso_cond} to hold in the perturbed state, then
\begin{align}
  \delta p = c_s^2(R,z) f^\prime \delta\rho,
\end{align}
where $f^\prime = df/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
obtain
\begin{align}\label{tbg_3d_2}
  T_\mathrm{BG} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\delta\rho\bm{\xi}^*\cdot\nabla c_s^2 +
    \xi_R^*\xi_z \left(\frac{\p\rho}{\p z}\frac{\p c_s^2}{\p R} -
      \frac{\p\rho}{\p R}\frac{\p c_s^2}{\p z}\right)\right],
\end{align}
where the equilibrium equations were used. 
At this point setting $\xi_z=0$ gives $T_\mathrm{BG}$ for
perturbations with no vertical motion, 
\begin{align}
  T_\mathrm{BG,2D} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left(\delta\rho \xi_R^*  \p_R c_s^2 \right),
\end{align}
and is equivalent to the 2D
expression, Eq. \ref{baroclinic_torque}, with $\delta\rho $ replaced
by $\delta \Sigma$ and $p=c_s^2\rho$. % The 2D expression is convenient
% for razor-thin discs or perturbations without vertical motion because
% one can measure $\delta\rho$ (or $\delta \Sigma)$ directly.

In fact, we can simplify Eq. \ref{tbg_3d_2} in the general case by
using $\delta\rho = - \rho\nabla\cdot\bm{\xi} -
\bm{\xi}\cdot\nabla\rho$, giving
\begin{align}\label{tbg_general}
  T_\mathrm{BG} = \frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\rho\left(\nabla\cdot\bm{\xi}\right)\bm{\xi}^*\cdot\nabla
  c_s^2\right].  
\end{align} 
For a barotropic fluid $p=p(\rho)$, the function $c_s^2$ can be
taken as constant (Eq. \ref{iso_cond}) for which $T_\mathrm{BG}$
vanishes. When there is a forced temperature gradient,
Eq. \ref{tbg_general} indicates there is angular momentum exchange
between compressible perturbations ($\nabla\cdot\bm{\xi}\neq0$) and
the background disc if there is motion along the temperature
gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).

% total effect needs to be integrated

% For the perturbations without vertical motion or those in razor-thin
% discs, the expression Eq. \ref{}

%eq baroclinic torque is more practical in 2d case because \delta
%sigma can be measured direction from simulations (xi needs to be
%calculated from velocity field, knowing the eigenfrequency)

### 2014-11-05 paper/discussion.tex
appendix and minor updates

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align}
  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\Sigma_m \left(\frac{\delta
        v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \Sigma_m \left(\frac{\delta
        v_{Rm}}{\sbar}\right)^*\right]
\end{align}
where we replaced the radial Lagrangian displacement by the Eulerian
radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
The linearised momentum equations give
\begin{align}
  D\delta v_{Rm} =& 
  \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
    + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
  &- \frac{2\ii
    m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
    \delta\Phi_m\right),
\end{align}
where $D\equiv \kappa^2 - \sbar^2$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
consistent with simulation results.

@@

% \subsection{Role of self-gravity and disc structure} 
% The background torque density $T_{BG}$ does not depend on
% self-gravity. 
% %and by definition, disc structure plays no role in the local description of linear modes 
% However, our numerical simulations show that the $m=1$ spiral is
% confined between two $Q$-barriers, where real solutions to the local
% dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
% structure. Thus, the role of disc structure and self-gravity is to
% allow a neutral mode to be set up and confined, which is then
% destabilised through the background torque density.

%talk about additional simulations at high Q
%spiral mode as external potential to wider disc (lp11)

>>>>

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align}
  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
        v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
        v_{Rm}}{\sbar}\right)^*\right]
\end{align}
where we replaced the radial Lagrangian displacement by the Eulerian
radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
The linearised momentum equations give
\begin{align}
  D\delta v_{Rm} =& 
  \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
    + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
  &- \frac{2\ii
    m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
    \delta\Phi_m\right),
\end{align}
where $D\equiv \kappa^2 - \sbar^2$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}\label{theoretical_rate}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
consistent with simulation results.

@@

\subsection{Role of self-gravity and disc structure} 
We note that the background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity, and we only considered a self-gravitating perturbation to
obtain an appropriate wavenumber $k$ needed to evaluate
Eq. \ref{theoretical_rate}. Thus, self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, the role of disc structure and self-gravity is to
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

%talk about additional simulations at high Q
%spiral mode as external potential to wider disc (lp11)
%interaction across co-rotation

### 2014-11-05 paper/model.tex
appendix and minor updates

The term $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided boundary fluxes vanish.

>>>>

The term $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momeumtum flux. We briefly outline the derivation of $T_\mathrm{BG}$
in Appendix 
\ref{tbg_deriv}.

### 2014-11-05 paper/paper.tex
appendix and minor updates

\maketitle
\begin{abstract} 
  We present numerical simulations of gravitational instability in
  astrophysical discs with an an imposed temperature profile, and in
  which only a sub-portion of it is self-gravitating. We conduct
  numerical hydrodynamic experiments in 2D using the FARGO code and in
  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
  development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
  outside the bulk of the spiral arm, resulting in a nearly stationary
  one-arm spiral pattern in the self-gravitating region. This one-arm
  spiral does not require the movement of 
  the central star, as found in previous studies. Instead, we find it
  to be associated with an imposed
  temperature gradient and the radial disc structure. 
  We apply results
  from existing linear theory and attribute the growth of this spiral
  to the non-conservation of angular momentum for linear perturbations
  in discs with a forced temperature gradient. This enables angular
  momentum exchange between the spiral and the background disc.

  % or confirm in 3D
  %, taken as a toy model of a 
  %protoplanetary disc containing a finite-size massive dead zone.
   % Theoretical models of circumstellar or protoplanetary discs indicate
  % the possible existence of `dead zones' --- regions in the disc
  % where mass accretion is greatly reduced because of inefficient
  % magneto-hydrodynamic turbulence that would otherwise enable 
  % accretion. It has been suggested that mass built-up in a dead zone
  % may lead to gravitational instability. However, what type of
  % gravitational instability is relevant in this context has not been
  % clarified.

  %
  % Our simulations demonstrate the 
  % importance of global gravitational interaction between a 
  % self-gravitating dead zone and the exterior, non-self-gravitating
  % parts of the disc. This non-local effect, due to disc structure, 
  % exists even when the total disc mass is below the threshold for
  % global effects found in
  % previous studies. 
\end{abstract}

@@

\appendix
%\input{appendix}
%\input{appendix2}

>>>>

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  systems we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region. This one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for linear perturbations in discs with a forced temperature
  gradient. This enables angular momentum exchange between the spiral
  and the background disc, and may lead to instability if the  
  interaction increases the amplitude of the linear perturbations.

  % or confirm in 3D
  %, taken as a toy model of a 
  %protoplanetary disc containing a finite-size massive dead zone.
   % Theoretical models of circumstellar or protoplanetary discs indicate
  % the possible existence of `dead zones' --- regions in the disc
  % where mass accretion is greatly reduced because of inefficient
  % magneto-hydrodynamic turbulence that would otherwise enable 
  % accretion. It has been suggested that mass built-up in a dead zone
  % may lead to gravitational instability. However, what type of
  % gravitational instability is relevant in this context has not been
  % clarified. 
  %
  % Our simulations demonstrate the 
  % importance of global gravitational interaction between a 
  % self-gravitating dead zone and the exterior, non-self-gravitating
  % parts of the disc. This non-local effect, due to disc structure, 
  % exists even when the total disc mass is below the threshold for
  % global effects found in
  % previous studies. 
\end{abstract}

@@

\appendix
\input{appendix}
%\input{appendix2}

### 2014-11-05 paper/results_3d.tex
appendix and minor updates

\section{Three-dimensional simulations}
In this section we review results from 3D simulations carried
out using the ZEUS-MP and PLUTO codes.

The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
dominated by an $m=1$ spiral.%  A Cartesian visualisation of the
% midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
The two codes give qualitatively similar results, but a longer PLUTO simulation was
needed to achieve similar mode amplitudes than ZEUS-MP. 
The amplitude of the spiral is smaller than the 2D case above,
possibily due to lower resolution and/or 3D self-gravity being weaker
than that in 2D.

>>>>

\section{Three-dimensional simulations}
In this section we briefly review results from 3D simulations carried 
out using the ZEUS-MP and PLUTO codes. The main purpose is to check
verify the above results with different numerical codes, and validate 
the 2D approximation.

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent 
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ spiral in the dead zone, which was found to dominate
the 2D simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=100P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
dominated by an $m=1$ spiral. %A Cartesian visualisation of the
% midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
The two codes give qualitatively similar results, but a longer PLUTO simulation was
needed to achieve similar mode amplitudes in ZEUS-MP.

%The amplitude of the spiral is smaller than the 2D case above,
%possibily due to lower resolution and/or 3D self-gravity being weaker
%than that in 2D.
